Analysis Script

### 2025-04-21 code/analysis-code/statisitical-analysis.qmd
Revert "Work"
This reverts commit 63ab86524d9b8d1ab4357a823b03d789a625d851.

Load needed packages. make sure they are installed.

```{r}
library(ggplot2) 
library(broom) 
library(here) 
library(glmnet)
library(MASS)
library(tidymodels)
library(dplyr)
library(rsample)
library(poissonreg)
library(parsnip)
library(future)
```

```{r}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")

#load data. 
data <- readRDS(data_location)

# NOTES TO SELF
# - change to QMD file (personally prefer)
# - work on other analysis (lag-time & quarantine fatigue? more research)
# - work on inclusion of county health ranking data or possibly not include at all? lots of mismatch counties between datasets
class(data$new_cases)
```

```{r}
# Intial Recipe setting
recipe <- recipe(new_cases ~ pop_density + transit_stations_percent_change_from_baseline + 
                     parks_percent_change_from_baseline + 
                     retail_and_recreation_percent_change_from_baseline + 
                     grocery_and_pharmacy_percent_change_from_baseline +
                     workplaces_percent_change_from_baseline + 
                     residential_percent_change_from_baseline +
                     health_factor_rank, data = data)

@@

First, I created a linear regression model as the baseline standard. This model performs very poorly as it has a low R-squared, indicating low predictive power. The model does mark all variables of the model to be significant.Additionally, the AIC is very extremely high.

```{r}
# Create linear regression model
lm_spec <- linear_reg() %>%
  set_engine("lm")

lm_wf <- workflow() %>%
  add_recipe(recipe)

lm_fit <- lm_wf %>%
  add_model(lm_spec) %>%
  fit(data = data)

lm_fit %>%
  extract_fit_parsnip() %>%
  tidy()

lm_fit %>%
  extract_fit_parsnip() %>%
  broom::glance()

lm_preds<- predict(lm_fit, new_data = data)

lm_metrics <- tibble(truth = data$new_cases, predicted = lm_preds$.pred) %>%
  metrics(truth = truth, estimate = predicted)
lm_metrics
```

I utilized Lasso here to hopefully bring down the RMSE and bring up the rsquared. It seems Lasso has no effect on model performance as the data lacks multicolinearity and does not have irrelevant predictors. 
```{r}
# Set method for LASSO
lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
  set_engine("glmnet")

# Insert recipe
lasso_wf <- workflow() %>%
  add_recipe(recipe)

lasso_fit <- lasso_wf %>%
  add_model(lasso_spec) %>%
  fit(data = data)

lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy()

lasso_fit %>%
  extract_fit_parsnip() %>%
  broom::glance()

lasso_preds<- predict(lasso_fit, new_data = data)

lasso_metrics <- tibble(truth = data$new_cases, predicted = lasso_preds$.pred) %>%
  metrics(truth = truth, estimate = predicted)
lasso_metrics
```

After the last two models performed poorly, the use of random forest seemed rational as this model can capture more complex relationships. 
```{r}

rngseed = 1234

forest_spec <- rand_forest(mode = "regression") %>%
  set_engine("ranger", seed = rngseed)

# Insert recipe
forest_wf <- workflow() %>%
  add_recipe(recipe)

# Create model with all predictors and random forest
forest_fit <- forest_wf %>%
 add_model(forest_spec) %>%
  fit(data = data)

forest_preds <- predict(forest_fit, new_data = data)
forest_preds <- tibble(truth = data$new_cases, predicted = forest_preds$.pred)
forest_rmse <- forest_preds %>%
  metrics(truth = truth, estimate = predicted)
print(forest_rmse)
```

```{r}
# Tuning forest
forest_spec <- rand_forest(
  mode = "regression", 
  trees = 300, 
  mtry = tune(), 
  min_n = tune()
) %>%
  set_engine("ranger")

# Create grid with set parameters
forest_grid <- grid_regular(
  mtry(range = c(1, 7)),  
  min_n(range = c(1, 21)),  
  levels = 7  
)

# Set seed
set.seed(1234)

# 5 fold crossvalidation repeated 5 times
# Note: Limitation of computing power so doing rolling window would not work
cv_folds <- vfold_cv(data, v = 5, repeats = 5)
plan(multisession, workers = 4)

# Tune the forest
# This step took 20-ish minutes
forest_tune_res <- tune_grid(
  forest_wf %>% add_model(forest_spec),
  resamples = cv_folds,  
  grid = forest_grid,
  metrics = metric_set(rmse, rsq),
  control = control_grid(parallel_over = "everything")
)

# Plot results
autoplot(forest_tune_res)

forest_tune_res %>%
  collect_metrics()

best_params <- select_best(forest_tune_res, metric = "rmse")
best_params

best_params$trees <- 500
```

```{r}
forest_spec <- rand_forest(
  mode = "regression",
  mtry = best_params$mtry,
  trees = best_params$trees,
  min_n = best_params$min_n
) %>%
  set_engine("ranger", seed = 1234)

forest_fit_final <- forest_wf %>%
  add_model(forest_spec) %>%
  fit(data = data)

# Get final predictions
forest_preds_final <- predict(forest_fit_final, new_data = data)
forest_preds_final <- tibble(truth = data$new_cases, predicted = forest_preds_final$.pred)

# Calculate performance metrics
forest_rmse <- forest_preds_final %>%
  metrics(truth = truth, estimate = predicted)

print(forest_rmse)
```

>>>>

Load needed packages
```{r}
library(ggplot2) 
library(broom) 
library(here) 
library(glmnet)
library(MASS)
library(tidymodels)
library(dplyr)
library(rsample)
library(tibble)
library(parsnip)
library(future)
library(vip)
library(lubridate)
library(patchwork)
```

Load in processed data and set eed
```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")

# load data 
data <- readRDS(data_location)

# set seed
rngseed = 1234
set.seed(rngseed)
```

Creation of lagged predictors
```{r}
# Creation of lagged cases and lagged mobility data
data <- data %>%
  group_by(county) %>%
  arrange(date) %>%
  mutate(
    lag_1 = lag(new_cases, 1),
    lag_7 = lag(new_cases, 7),
    lag_14 = lag(new_cases, 14),
    retail_lag7 = lag(retail_and_recreation_percent_change_from_baseline, 7),
    grocery_lag7 = lag(grocery_and_pharmacy_percent_change_from_baseline, 7),
    workplace_lag7 = lag(workplaces_percent_change_from_baseline, 7),
    residential_lag7 = lag(residential_percent_change_from_baseline, 7)
  ) %>%
  ungroup()

# remove NAs as result of lagging
data <- data %>% drop_na()
```

Implementation of rolling window cross-validation
```{r}
# Arrange by date
data <- data %>%
  arrange(date, county)

# Define testing set as last 30 days for 44 counties
test_days <- 30
last_date <- max(data$date)
test_start_date <- last_date - days(test_days - 1)

# Define train and test split through dates
train_data <- data %>% filter(date < test_start_date)
test_data  <- data %>% filter(date >= test_start_date)

# Create rolling window folds for cross-validation 
rolling_folds <- rolling_origin(
  data = train_data,
  initial = 120 * 44,  # ~4 months of training
  assess  = 30 * 44,   # ~1 month of validation
  skip    = 15 * 44,   # roll forward 15 days
  cumulative = FALSE
)

# Check date ranges for each fold
train_ranges <- purrr::map(rolling_folds$splits, ~range(pull(training(.x), date)))

# Print the ranges for each fold
train_ranges

# View fold summary
fold_sizes <- purrr::map_dfr(rolling_folds$splits, function(split) {
  tibble(
    train_n = nrow(training(split)),
    test_n  = nrow(testing(split))
  )
})

print(fold_sizes)

# Check date ranges
range(train_data$date)
range(test_data$date)

@@

Recipe setting for modeling
```{r}
# Recipe for baseline model
recipe_baseline <- recipe(new_cases ~ pop_density + lag_1 + lag_7 + lag_14, data = data)

# Recipe with all predictors
recipe <- recipe(new_cases ~ pop_density +
                  retail_and_recreation_percent_change_from_baseline +
                   grocery_and_pharmacy_percent_change_from_baseline + 
                   workplaces_percent_change_from_baseline +
                   residential_percent_change_from_baseline +
                     lag_1 + lag_7 + lag_14, data = data)

# Recipe with lagged versions of mobility
recipe_lag <- recipe(new_cases ~ pop_density + retail_lag7 + grocery_lag7 + residential_lag7 + workplace_lag7 + lag_1 + lag_7 + lag_14, data = data)
```

Modeling

The basic workflow is to fit a baseline model, fit the mobility model, and then fit the lagged model.

First model are linear regression models.
This code chunk fits a baseline linear regression model.
```{r}
# Define the Lasso model with tuning parameters
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")

# Create the Lasso workflow with baseline recipe
base_tuned_lasso_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(lasso_spec)

# Define a grid for tuning the penalty
lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 30)

# Tune the Lasso model using rolling origin cv
base_tuned_lasso_res <- tune_grid(
  base_tuned_lasso_wf,
  resamples = rolling_folds,
  grid = lasso_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything", save_pred = TRUE)
)

# View tuning results
autoplot(base_tuned_lasso_res)
base_tuned_lasso_metrics <- collect_metrics(base_tuned_lasso_res)
print(base_tuned_lasso_metrics)

# Select best penalty based on R-squared
base_tuned_lasso_best <- select_best(base_tuned_lasso_res, metric = "rsq")
print(base_tuned_lasso_best)

# Finalize the Lasso model with the selected penalty
base_final_lasso_spec <- linear_reg(
  penalty = base_tuned_lasso_best$penalty,
  mixture = 1
) %>%
  set_engine("glmnet")

# Final workflow with tuned Lasso model
base_final_lasso_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(base_final_lasso_spec)

# Fit the final Lasso model on training data
base_final_lasso_fit <- base_final_lasso_wf %>%
  fit(data = train_data)

# Evaluate the final Lasso model with rolling CV
base_final_lasso_res <- fit_resamples(
  base_final_lasso_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# View resampling performance for the final Lasso model
collect_metrics(base_final_lasso_res)

# Predict on test data using the final Lasso model
base_final_lasso_test_preds <- predict(base_final_lasso_fit, new_data = test_data)

# Combine predictions with actual values
base_final_lasso_test_results <- test_data %>%
  select(new_cases) %>%
  mutate(predicted = base_final_lasso_test_preds$.pred)

# Compute test metrics for the final Lasso model
base_final_lasso_test_metrics <- base_final_lasso_test_results %>%
  metrics(truth = new_cases, estimate = predicted)

# View the test set performance for the final Lasso model
print(base_final_lasso_test_metrics)

# View variable importance for the Lasso model
vip(base_final_lasso_fit$fit$fit)

This code chunk fits a linear regression model with mobility predictors.
```{r}
# Create the Lasso workflow
lasso_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(lasso_spec)

# Tune the Lasso model using time-blocked CV
lasso_res <- tune_grid(
  lasso_wf,
  resamples = rolling_folds,
  grid = lasso_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(save_pred = TRUE)
)

# View tuning results
autoplot(lasso_res)
lasso_metrics <- collect_metrics(lasso_res)
print(lasso_metrics)

# Select best penalty by R-squared
lasso_best <- select_best(lasso_res, metric = "rsq")
print(lasso_best)

# Finalize the Lasso model with selected lambda
final_lasso_spec <- linear_reg(
  penalty = lasso_best$penalty,
  mixture = 1
) %>%
  set_engine("glmnet")

# Final workflow with tuned Lasso model
final_lasso_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(final_lasso_spec)

# Fit final model on training data
final_lasso_fit <- final_lasso_wf %>%
  fit(data = train_data)

# Evaluate final model with rolling CV
final_lasso_res <- fit_resamples(
  final_lasso_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# View resampling performance
collect_metrics(final_lasso_res)

# Predict on test data
final_lasso_test_preds <- predict(final_lasso_fit, new_data = test_data)

# Combine predictions with actuals
final_lasso_test_results <- test_data %>%
  select(new_cases) %>%
  mutate(predicted = final_lasso_test_preds$.pred)

# Compute test metrics
final_lasso_test_metrics <- final_lasso_test_results %>%
  metrics(truth = new_cases, estimate = predicted)

# View test set performance
print(final_lasso_test_metrics)

# View variable importance
vip(final_lasso_fit$fit$fit)
```

```{r}
# Create the Lasso workflow
lag_lasso_wf <- workflow() %>%
  add_recipe(recipe_lag) %>%
  add_model(lasso_spec)

# Tune the Lasso model using time-blocked CV
lag_lasso_res <- tune_grid(
  lag_lasso_wf,
  resamples = rolling_folds,
  grid = lasso_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(save_pred = TRUE)
)

# View tuning results
autoplot(lag_lasso_res)
lag_lasso_metrics <- collect_metrics(lag_lasso_res)
print(lag_lasso_metrics)

# Select best penalty by R-squared
lag_lasso_best <- select_best(lag_lasso_res, metric = "rsq")
print(lag_lasso_best)

# Finalize the Lasso model with selected lambda
lag_final_lasso_spec <- linear_reg(
  penalty = lag_lasso_best$penalty,
  mixture = 1
) %>%
  set_engine("glmnet")

# Final workflow with tuned Lasso model
lag_final_lasso_wf <- workflow() %>%
  add_recipe(recipe_lag) %>%
  add_model(lag_final_lasso_spec)

# Fit final model on training data
lag_final_lasso_fit <- lag_final_lasso_wf %>%
  fit(data = train_data)

# Evaluate final model with rolling CV
lag_final_lasso_res <- fit_resamples(
  lag_final_lasso_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# View resampling performance
collect_metrics(lag_final_lasso_res)

# Predict on test data
lag_final_lasso_test_preds <- predict(lag_final_lasso_fit, new_data = test_data)

# Combine predictions with actuals
lag_final_lasso_test_results <- test_data %>%
  select(new_cases) %>%
  mutate(predicted = lag_final_lasso_test_preds$.pred)

# Compute test metrics
lag_final_lasso_test_metrics <- lag_final_lasso_test_results %>%
  metrics(truth = new_cases, estimate = predicted)

# View test set performance
print(lag_final_lasso_test_metrics)

# View variable importance
vip(lag_final_lasso_fit$fit$fit)

```
```{r}
# plot baseline lasso model prediction vs actual
plot_base_lasso <- base_final_lasso_test_results %>%
  ggplot(aes(x = new_cases, y = predicted)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Base Lasso",
    x = "Actual Cases", y = "Predicted Cases"
  ) +
  theme_bw()

# plot mobility lasso model prediction vs actual
plot_lasso <- final_lasso_test_results %>%
  ggplot(aes(x = new_cases, y = predicted)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Lasso",
    x = "Actual Cases", y = "Predicted Cases"
  ) +
  theme_bw()

# plot lagged lasso model prediction vs actual
plot_lag_lasso <- lag_final_lasso_test_results %>%
  ggplot(aes(x = new_cases, y = predicted)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Lag Lasso",
    x = "Actual Cases", y = "Predicted Cases"
  ) +
  theme_bw()

lasso_graphs <- plot_base_lasso + plot_lasso + plot_lag_lasso +
  plot_layout(ncol = 3, widths = c(1.5, 1.5, 1.5)) &
  theme(
    plot.margin = margin(10, 10, 10, 10),
    plot.title = element_text(size = 14, hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(lasso_graphs)

figure_file = here("results", "figures", "lasso_graphs.png")
ggsave(
  filename = figure_file,
  plot = lasso_graphs,
  width = 16, height = 5, dpi = 300
)

Second, random forest models are tuned, trained, and evaluated. Best models are chosen through R-squared.

This code chunk fits and tunes a baseline RF model with no mobility predictors.
```{r}
# Define the Random Forest model with tuning parameters
forest_spec <- rand_forest(
  mode = "regression", 
  mtry = tune(),           
  min_n = tune(),          
  trees = 500              
) %>%
  set_engine("ranger", seed = rngseed)

# Create grid of hyperparameters for tuning
forest_grid <- grid_regular(
  mtry(range = c(1, 4)),  
  min_n(range = c(1, 50)), 
  levels = 10
)

# Create baseline workflow with baseline recipe
base_tuned_rf_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(forest_spec)

# Tune the Random Forest model using time-blocked CV (folds)
base_tuned_rf_res <- tune_grid(
  base_tuned_rf_wf,
  resamples = rolling_folds,                     
  grid = forest_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# View tuning results
autoplot(base_tuned_rf_res)
base_tune_rf_metrics <- collect_metrics(base_tuned_rf_res)
print(base_tune_rf_metrics)

# Select best hyperparameters by R-squared
base_tuned_rf_best <- select_best(base_tuned_rf_res, metric = "rsq")

# Finalize model with best parameters
base_final_rf_spec <- rand_forest(
  mode = "regression",
  mtry = base_tuned_rf_best$mtry,
  min_n = base_tuned_rf_best$min_n,
  trees = 500
) %>%
  set_engine("ranger", seed = rngseed)

# Final workflow with tuned model
base_final_rf_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(base_final_rf_spec)

# Fit final model on training data
base_final_rf_fit <- base_final_rf_wf %>%
  fit(data = train_data)

# Evaluate final model with rolling CV
base_final_rf_res <- fit_resamples(
  base_final_rf_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# View resampling performance
collect_metrics(base_final_rf_res)

# Predict on test data
base_final_rf_test_preds <- predict(base_final_rf_fit, new_data = test_data)

# Combine predictions with actuals
base_final_rf_test_results <- test_data %>%
  select(new_cases) %>%
  mutate(predicted = base_final_rf_test_preds$.pred)

# Compute test metrics
base_final_rf_test_metrics <- base_final_rf_test_results %>%
  metrics(truth = new_cases, estimate = predicted)

# View test set performance
print(base_final_rf_test_metrics)

This code chunk fits and tunes a RF model with mobility predictors.
```{r}
# Create workflow with baseline recipe
tuned_rf_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(forest_spec)

# Tune the Random Forest model using time-blocked CV (folds)
rf_tune_res <- tune_grid(
  tuned_rf_wf,
  resamples = rolling_folds,                     
  grid = forest_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# View tuning results
autoplot(rf_tune_res)
rf_tune_metrics <- collect_metrics(rf_tune_res)
print(rf_tune_metrics)

# Select best hyperparameters by R-squared
rf_best <- select_best(rf_tune_res, metric = "rsq")

# Finalize model with best parameters
final_rf_spec <- rand_forest(
  mode = "regression",
  mtry = rf_best$mtry,
  min_n = rf_best$min_n,
  trees = 500
) %>%
  set_engine("ranger", seed = rngseed)

# Final workflow with tuned model
final_rf_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(final_rf_spec)

# Fit final model on training data
final_rf_fit <- final_rf_wf %>%
  fit(data = train_data)

# Evaluate final model with rolling CV
final_rf_res <- fit_resamples(
  final_rf_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# View resampling performance
collect_metrics(final_rf_res)

# Predict on test data
final_rf_test_preds <- predict(final_rf_fit, new_data = test_data)

# Combine predictions with actuals
final_rf_test_results <- test_data %>%
  select(new_cases) %>%
  mutate(predicted = final_rf_test_preds$.pred)

# Compute test metrics
final_rf_test_metrics <- final_rf_test_results %>%
  metrics(truth = new_cases, estimate = predicted)

# View test set performance
print(final_rf_test_metrics)
```

This code chunk fits and tunes a RF model with lagged mobility predictors.
```{r}
# Create lag workflow with baseline recipe
lag_tuned_rf_wf <- workflow() %>%
  add_recipe(recipe_lag) %>%
  add_model(forest_spec)

# Tune the Random Forest model using time-blocked CV (folds)
lag_tuned_rf_res <- tune_grid(
  lag_tuned_rf_wf,
  resamples = rolling_folds,                     
  grid = forest_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# View tuning results
autoplot(lag_tuned_rf_res)
lag_tune_rf_metrics <- collect_metrics(lag_tuned_rf_res)
print(lag_tune_rf_metrics)

# Select best hyperparameters by R-squared
lag_tuned_rf_best <- select_best(lag_tuned_rf_res, metric = "rsq")

# Finalize model with best parameters
lag_final_rf_spec <- rand_forest(
  mode = "regression",
  mtry = lag_tuned_rf_best$mtry,
  min_n = lag_tuned_rf_best$min_n,
  trees = 500
) %>%
  set_engine("ranger", seed = rngseed)

# Final workflow with tuned model
lag_final_rf_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(lag_final_rf_spec)

# Fit final model on training data
lag_final_rf_fit <- lag_final_rf_wf %>%
  fit(data = train_data)

# Evaluate final model with rolling CV
lag_final_rf_res <- fit_resamples(
  lag_final_rf_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# View resampling performance
collect_metrics(lag_final_rf_res)

# Predict on test data
lag_final_rf_test_preds <- predict(lag_final_rf_fit, new_data = test_data)

# Combine predictions with actuals
lag_final_rf_test_results <- test_data %>%
  select(new_cases) %>%
  mutate(predicted = lag_final_rf_test_preds$.pred)

# Compute test metrics
lag_final_rf_test_metrics <- lag_final_rf_test_results %>%
  metrics(truth = new_cases, estimate = predicted)

# View test set performance
print(lag_final_rf_test_metrics)

```{r}
# plot baseline rf model prediction vs actual
plot_base_rf <- base_final_rf_test_results %>%
  ggplot(aes(x = new_cases, y = predicted)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Base RF",
    x = "Actual Cases", y = "Predicted Cases"
  ) +
  theme_bw()

# plot mobility rf model prediction vs actual
plot_rf <- final_rf_test_results %>%
  ggplot(aes(x = new_cases, y = predicted)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "RF with Mobility",
    x = "Actual Cases", y = "Predicted Cases"
  ) +
  theme_bw()

# plot lagged rf model prediction vs actual
plot_lag_rf <- lag_final_rf_test_results %>%
  ggplot(aes(x = new_cases, y = predicted)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "RF with Lagged Mobility",
    x = "Actual Cases", y = "Predicted Cases"
  ) +
  theme_bw()

rf_graphs <- plot_base_rf + plot_rf + plot_lag_rf +
  plot_layout(ncol = 3, widths = c(1.5, 1.5, 1.5)) &
  theme(
    plot.margin = margin(10, 10, 10, 10),
    plot.title = element_text(size = 14, hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(rf_graphs)

figure_file = here("results", "figures", "rf_graphs.png")
ggsave(
  filename = figure_file,
  plot = rf_graphs,
  width = 16, height = 5, dpi = 300
)
```

Third, XGBoost models are tuned, trained, and evaluated. Best models are chosen through R-squared.

This code chunk fits and tunes an baseline XGBoost model with no mobility predictors.
```{r}
# Define the XGBoost model specification (baseline version)
boost_spec <- boost_tree(
  trees = 500,
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Build workflow with baseline recipe and model
base_boost_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(boost_spec)

# Tune XGBoost model using time-blocked CV (folds)
base_boost_tune_res <- tune_grid(
  base_boost_wf,
  resamples = rolling_folds,
  grid = 30,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# View tuning results and print best model metrics
base_boost_tune_res %>%
  collect_metrics() %>%
  print()

base_best_params <- select_best(base_boost_tune_res, metric = "rsq")
print(base_best_params)

# Final XGBoost model with best hyperparameters
base_final_boost_spec <- boost_tree(
  trees = 500,
  tree_depth = base_best_params$tree_depth,
  learn_rate = base_best_params$learn_rate,
  loss_reduction = base_best_params$loss_reduction
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Final model workflow
base_final_boost_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(base_final_boost_spec)

# Fit the final model
base_final_boost_fit <- base_final_boost_wf %>%
  fit(data = train_data)

# Predict on test data and compute performance metrics
base_final_boost_preds <- predict(base_final_boost_fit, new_data = test_data)

base_final_boost_results <- tibble(
  truth = test_data$new_cases,
  predicted = base_final_boost_preds$.pred
)

base_final_boost_metrics <- base_final_boost_results %>%
  metrics(truth = truth, estimate = predicted)

# Print final performance metrics
print(base_final_boost_metrics)
```

This code chunk fits and tunes an XGBoost model with mobility predictors.
```{r}
# Build workflow with recipe and model
boost_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(boost_spec)

# Tune XGBoost model using time-blocked CV (folds)
boost_tune_res <- tune_grid(
  boost_wf,
  resamples = rolling_folds,
  grid = 30,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# View tuning results and print best model metrics
boost_tune_res %>%
  collect_metrics() %>%
  print()

best_params <- select_best(boost_tune_res, metric = "rsq")
print(best_params)

# Final XGBoost model with best hyperparameters
final_boost_spec <- boost_tree(
  trees = 500,
  tree_depth = best_params$tree_depth,
  learn_rate = best_params$learn_rate,
  loss_reduction = best_params$loss_reduction
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Final model workflow
final_boost_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(final_boost_spec)

# Fit the final model
final_boost_fit <- final_boost_wf %>%
  fit(data = train_data)

# Predict on test data and compute performance metrics
final_boost_preds <- predict(final_boost_fit, new_data = test_data)

final_boost_results <- tibble(
  truth = test_data$new_cases,
  predicted = final_boost_preds$.pred
)

final_boost_metrics <- final_boost_results %>%
  metrics(truth = truth, estimate = predicted)

# Print final performance metrics
print(final_boost_metrics)

This code chunk fits and tunes an XGBoost model with lagged mobility predictors.
```{r}
# Build workflow with baseline recipe and model
lag_boost_wf <- workflow() %>%
  add_recipe(recipe_lag) %>%
  add_model(boost_spec)

# Tune XGBoost model using time-blocked CV (folds)
lag_boost_tune_res <- tune_grid(
  lag_boost_wf,
  resamples = rolling_folds,
  grid = 30,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# View tuning results and print best model metrics
lag_boost_tune_res %>%
  collect_metrics() %>%
  print()

lag_best_params <- select_best(lag_boost_tune_res, metric = "rsq")
print(lag_best_params)

# Final XGBoost model with best hyperparameters
lag_final_boost_spec <- boost_tree(
  trees = 500,
  tree_depth = lag_best_params$tree_depth,
  learn_rate = lag_best_params$learn_rate,
  loss_reduction = lag_best_params$loss_reduction
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Final model workflow
lag_final_boost_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(lag_final_boost_spec)

# Fit the final model
lag_final_boost_fit <- lag_final_boost_wf %>%
  fit(data = train_data)

# Predict on test data and compute performance metrics
lag_final_boost_preds <- predict(lag_final_boost_fit, new_data = test_data)

lag_final_boost_results <- tibble(
  truth = test_data$new_cases,
  predicted = lag_final_boost_preds$.pred
)

lag_final_boost_metrics <- lag_final_boost_results %>%
  metrics(truth = truth, estimate = predicted)

# Print final performance metrics
print(lag_final_boost_metrics)
```

```{r}
# plot baseline boost model prediction vs. actual
plot_base_boost <- base_final_boost_results %>%
  ggplot(aes(x = truth, y = predicted)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Base Boost",
    x = "Actual Cases", y = "Predicted Cases"
  ) +
  theme_bw()

# plot mobility boost model prediction vs actual
plot_boost <- final_boost_results %>%
  ggplot(aes(x = truth, y = predicted)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Boost with Mobility",
    x = "Actual Cases", y = "Predicted Cases"
  ) +
  theme_bw()

# plot lagged boost model prediction vs actual
plot_lag_boost <- lag_final_boost_results %>%
  ggplot(aes(x = truth, y = predicted)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Boost with Lagged Mobility",
    x = "Actual Cases", y = "Predicted Cases"
  ) +
  theme_bw()

boost_graphs <- plot_base_boost + plot_boost + plot_lag_boost +
  plot_layout(ncol = 3, widths = c(1.5, 1.5, 1.5)) &
  theme(
    plot.margin = margin(10, 10, 10, 10),
    plot.title = element_text(size = 14, hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(boost_graphs)

figure_file = here("results", "figures", "boost_graphs.png")
ggsave(
  filename = figure_file,
  plot = boost_graphs,
  width = 16, height = 5, dpi = 300
)

This code chunk merges the results of the best models into a data frame.
```{r}
base_lasso_metrics <- base_final_lasso_test_metrics %>% 
  mutate(model = "LASSO (Baseline)") %>%
  rename(mean = .estimate)

lasso_metrics <- final_lasso_test_metrics %>% 
  mutate(model = "LASSO") %>%
  rename(mean = .estimate)

lag_lasso_metrics <- lag_final_lasso_test_metrics %>% 
  mutate(model = "LASSO (Lagged Mobility)") %>%
  rename(mean = .estimate)

final_boost_metrics <- final_boost_metrics %>%
  mutate(model = "XGBoost") %>%
  rename(mean = .estimate)

final_forest_metrics <- final_rf_test_metrics %>%
  mutate(model = "Random Forest") %>%
  rename(mean = .estimate)

base_final_boost_metrics <- base_final_boost_metrics %>%
  mutate(model = "XGBoost (Baseline)") %>%
  rename(mean = .estimate)

base_final_forest_metrics <- base_final_rf_test_metrics %>%
  mutate(model = "Random Forest (Baseline)") %>%
  rename(mean = .estimate)

lag_boost_metrics <- lag_final_boost_metrics %>%
  mutate(model = "XGBoost (Lagged Mobility)") %>%
  rename(mean = .estimate)

lag_forest_metrics <- lag_final_rf_test_metrics %>%
  mutate(model = "Random Forest (Lagged Mobility)") %>%
  rename(mean = .estimate)

model_metrics <- bind_rows(base_lasso_metrics, lasso_metrics, lag_lasso_metrics, base_final_boost_metrics, final_boost_metrics, lag_boost_metrics, base_final_forest_metrics, final_forest_metrics, lag_forest_metrics)

model_metrics <- model_metrics %>%
  select(-.estimator) %>%
  pivot_wider(names_from = '.metric', values_from = mean)
```

This code chunk creates a short summary table of the results from the model_metrics data frame.
```{r}
library(gt)

model_table <- model_metrics %>%
  gt() %>%
  tab_header(
    title = "Model Performance Metrics",
    subtitle = "Best values are highlighted per metric"
  ) %>%
  fmt_number(columns = c(rmse, mae), decimals = 1) %>%
  fmt_number(columns = rsq, decimals = 3) %>%
  cols_label(
    model = "Model",
    rmse = "RMSE",
    rsq = "R²",
    mae = "MAE"
  ) %>%

  # Highlight lowest RMSE
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(
      columns = rmse,
      rows = rmse == min(rmse, na.rm = TRUE)
    )
  ) %>%

  # Highlight lowest MAE
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(
      columns = mae,
      rows = mae == min(mae, na.rm = TRUE)
    )
  ) %>%

  # Highlight highest R²
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(
      columns = rsq,
      rows = rsq == max(rsq, na.rm = TRUE)
    )
  ) %>%
